---
title: "Class 11: Protein Structure Prediction with Alphafold"
author: "Cyrus Shabahang (PID: A19145663)"
format: pdf
toc: TRUE
---

## Background 

In this hands-on session we will utilize AlphaFold to predict protein structure from sequence (Jumper et al. 2021).

Without the aid of such approaches, it can take years of expensive laboratory work to determine the structure of just one protein. With AlphaFold we can now accurately compute a typical protein structure in as little as ten minutes.

The PDB database (the main repository of experimental structures) only has **~250 thousand** structures (we saw this in the last lab). The main protein sequence database has over **200 million** sequences! Only 0.125% of known sequences have a known structure - this is called the "structure knowledge gap".

```{r}
(250000 / 200000000) *100
```

- Structures are much harder to determine than sequences
- They are expensive (on average ~$1 million each)
- They take on average 3-5 years to solve! 

# EBI AlphaFold Database 

- The EBI has a database of pre-computed AlphaFold (AF) models called AFDB. 
- This is constantly growing and can be useful to check before running AF ourselves. 

## Running Alphafold

We can download and run locally (on our own computers) but we need a GPU. Or we can use "cloud" computing to run this on someone elses computers. 

We will use ColabFold <https://github.com/sokrypton/ColabFold> 

We previously found there was no AFDB entry for our HIV sequence: 

```
>HIV-Pr-Dimer
PQITLWQRPLVTIKIGGQLKEALLDTGADDTVLEEMSLPGRWKPKMIGGIGGFIKVRQYD
QILIEICGHKAIGTVLVGPTPVNIIGRNLLTQIGCTLNF:PQITLWQRPLVTIKIGGQLK
EALLDTGADDTVLEEMSLPGRWKPKMIGGIGGFIKVRQYDQILIEICGHKAIGTVLVGPT
PVNIIGRNLLTQIGCTLNF
```
Here we will use AlphaFold2_mmseqs2 

After I ran this in the online GPU resource, ColabFold get back a downloadable zip file that contained five ranked structure models in PDB format, MSA alignment files, and JSON files. 

## Interpreting Results 

I unzipped the downloadable files and looked in Mol to see a visualization and also in Bio3D. 

## Structural analysis in R with Bio3D

```{r}
library(bio3d)
library(jsonlite)
library(pheatmap)

results_dir <- "hivpr_23119/"
list.files(results_dir)[1:20]
```
```{r}
library(bio3d)

pdb_files <- list.files(path = results_dir, 
                        pattern = "\\.pdb$",
                        full.names = TRUE)
basename(pdb_files)
length(pdb_files)

pdbs <- pdbaln(pdb_files, fit = TRUE, exefile = "msa")
pdbs
```
RMSD (heatmap)

```{r}
library(pheatmap)

rd <- rmsd(pdbs, fit = TRUE)
range(rd)

colnames(rd) <- paste0("m", 1:nrow(rd))
rownames(rd) <- paste0("m", 1:nrow(rd))

pheatmap(rd)

```

From our RMSD range, we can see that certain models were very similar while others were very different from each other. 

```{r}

pdb_ref <- read.pdb("1hsg")   

plotb3(pdbs$b[1,], typ="l", lwd=2, sse=pdb_ref,
       ylab="pLDDT (B-factor)", xlab="Residue Position")

lines(pdbs$b[2,], col="red")
lines(pdbs$b[3,], col="blue")
lines(pdbs$b[4,], col="darkgreen")
lines(pdbs$b[5,], col="orange")

abline(v = 99, col = "gray")


```



```{r}
library(jsonlite)

pae_files <- list.files(path = results_dir,
                        pattern = "scores_rank_.*\\.json$",
                        full.names = TRUE)
basename(pae_files)
pae1 <- read_json(pae_files[1], simplifyVector = TRUE)
pae5 <- read_json(pae_files[5], simplifyVector = TRUE)
pae1$max_pae
pae5$max_pae
plot.dmat(pae1$pae, xlab="Residue i", ylab="Residue j",
          grid.col="black", zlim=c(0,30))
plot.dmat(pae5$pae, xlab="Residue i", ylab="Residue j",
          grid.col="black", zlim=c(0,30))

```

.a3m file:

```{r}
aln_file <- list.files(path = results_dir,
                       pattern = "\\.a3m$",
                       full.names = TRUE)

aln <- read.fasta(aln_file[1], to.upper = TRUE)
dim(aln$ali)
sim <- conserv(aln)
plotb3(sim[1:99], ylab="Conservation Score", xlab="Residue Position")

```
Core fit + RMSF

```{r}
core <- core.find(pdbs)
core.inds <- print(core, vol = 0.5)

xyz <- pdbfit(pdbs, core.inds, outpath = "corefit_structures")

rf <- rmsf(xyz)
plotb3(rf, sse = pdb_ref, ylab="RMSF (Å)", xlab="Residue Position")
abline(v = 99, col = "gray")

```
Mapping conservation onto a PDB for Mol

```{r}
m1.pdb <- read.pdb(pdb_files[1])

occ <- vec2resno(c(sim[1:99], sim[1:99]), m1.pdb$atom$resno)

write.pdb(m1.pdb, o = occ, file = "m1_conserv.pdb")

```



