---
title: "Class 12: Genome Informatics Lab Q13&Q14"
author: "Cyrus Shabahang (PID:19145663)"
format: pdf
---

> Q13 - Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
url_w19 <- "https://bioboot.github.io/bggn213_W19/class-material/rs8067378_ENSG00000172057.6.txt"
dat <- read.table(url_w19, header=TRUE, stringsAsFactors=FALSE)
head(dat)
str(dat)
colnames(dat)
```

```{r}
n_by_geno <- table(dat$geno)
n_by_geno
med_by_geno <- tapply(dat$exp, dat$geno, median)
med_by_geno

summary_tbl <- data.frame(
  genotype = names(n_by_geno),
  n = as.vector(n_by_geno), 
  median_exp = as.vector(med_by_geno)
  
)
summary_tbl
```
The sample sizes are: A/A: 108, A/G: 233, and G/G: 121. The median expression values are 31.24847 (A/A), 25.06486 (A/G), and 20.07363 (G/G). 


> Q14 -  Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
boxplot(exp ~ geno, data=dat,
        xlab = "rs8067378 genotype", 
        ylab = "ORMDL3 expression",
        main = "ORMDL3 expression vs rs8067378 genotype")
```

When the A/A box has a higher median than G/G, the A allele has a corresponding higher ORMDL3 expression. A/A has the highest expression, A/G is intermediate, and G/G has the lowest expression. If the G/G box has a higher median than A/A, then the G allele has a higher ORMDL3 expression. The SNP does affect the expression of ORMDL3.  