---
title: "Class 18: Pertussis Mini Project"
author: "Cyrus Shabahang (PID: 19145663)"
format: html
---

##Background 

Pertussis (a.k.a whooping cough) is a common lung infection caused by the bacteria B. Pertussis. 

This can infect all ages but is most severe for those under 1 year of age. 

The CDC track the number of [reported cases in the US](https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html) 

We can "scrape" this data with the **datapasta** package. 

```{r}
cdc <- data.frame(
  year = c(
    1922L,
    1923L,
    1924L,
    1925L,
    1926L,
    1927L,
    1928L,
    1929L,
    1930L,
    1931L,
    1932L,
    1933L,
    1934L,
    1935L,
    1936L,
    1937L,
    1938L,
    1939L,
    1940L,
    1941L,
    1942L,
    1943L,
    1944L,
    1945L,
    1946L,
    1947L,
    1948L,
    1949L,
    1950L,
    1951L,
    1952L,
    1953L,
    1954L,
    1955L,
    1956L,
    1957L,
    1958L,
    1959L,
    1960L,
    1961L,
    1962L,
    1963L,
    1964L,
    1965L,
    1966L,
    1967L,
    1968L,
    1969L,
    1970L,
    1971L,
    1972L,
    1973L,
    1974L,
    1975L,
    1976L,
    1977L,
    1978L,
    1979L,
    1980L,
    1981L,
    1982L,
    1983L,
    1984L,
    1985L,
    1986L,
    1987L,
    1988L,
    1989L,
    1990L,
    1991L,
    1992L,
    1993L,
    1994L,
    1995L,
    1996L,
    1997L,
    1998L,
    1999L,
    2000L,
    2001L,
    2002L,
    2003L,
    2004L,
    2005L,
    2006L,
    2007L,
    2008L,
    2009L,
    2010L,
    2011L,
    2012L,
    2013L,
    2014L,
    2015L,
    2016L,
    2017L,
    2018L,
    2019L,
    2020L,
    2021L,
    2022L,
    2023L
  ), cases = c(
    107473,
    164191,
    165418,
    152003,
    202210,
    181411,
    161799,
    197371,
    166914,
    172559,
    215343,
    179135,
    265269,
    180518,
    147237,
    214652,
    227319,
    103188,
    183866,
    222202,
    191383,
    191890,
    109873,
    133792,
    109860,
    156517,
    74715,
    69479,
    120718,
    68687,
    45030,
    37129,
    60886,
    62786,
    31732,
    28295,
    32148,
    40005,
    14809,
    11468,
    17749,
    17135,
    13005,
    6799,
    7717,
    9718,
    4810,
    3285,
    4249,
    3036,
    3287,
    1759,
    2402,
    1738,
    1010,
    2177,
    2063,
    1623,
    1730,
    1248,
    1895,
    2463,
    2276,
    3589,
    4195,
    2823,
    3450,
    4157,
    4570,
    2719,
    4083,
    6586,
    4617,
    5137,
    7796,
    6564,
    7405,
    7298,
    7867,
    7580,
    9771,
    11647,
    25827,
    25616,
    15632,
    10454,
    13278,
    16858,
    27550,
    18719,
    48277,
    28639,
    32971,
    20762,
    17972,
    18975,
    15609,
    18617,
    6124,
    2116,
    3044,
    7063
  )
)
```

> Q1. Make a plot of `year` vs `cases`


```{r}
library(ggplot2)

ggplot(cdc) +
  aes(year, cases) +
  geom_point() +
  geom_line()
```

> Q2. Add some major milestones including the first wP vaccine rollout (1946), the switch to the newer aP vaccine (1996). 



```{r}
ggplot(cdc) +
  aes(year, cases) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 1946, col = "blue", lty = 2) +
  geom_vline(xintercept = 1996, col = "red", lty = 2) +
  geom_vline(xintercept = 2020, col = "darkgreen", lty = 2) 
```



> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

There were high case numbers in the pre 1940s, then came the introduction of the aP vaccine. This dramatically reduced the number of cases. Differing from the wP vaccine, the aP vaccine needs a booster vaccine after around 10 years. Around the 2000s we can see a rise in cases, most likely due to people not getting booster vaccines. Around 2020, we can see cases drop due to quarantine for COVID-19, but rise again likely due to anti-vaxxers from the fear of the COVID-19 vaccine. The aP induced protection wanes faster than wP.

**Why is this vaccine-preventable disease on the upswing?** To answer this question we need to investigate the mechanisms underlying waning protection against pertussis. This requires evaluation of pertussis-specific immune responses over time in wP and aP vaccinated individuals.

## CMI-PB project
[Computational Models of Immunity](https://www.cmi-pb.org/) - Pertussis Boost project aims to provide the scientific comunity with this very information. 

They make their data available via JSON format APIs. We can read this in R with the `read_json()` function from the **jsonlite** package: 


```{r}
library(jsonlite)

subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 

head(subject)
```

>Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```


There are 87 aP and 85 wP infancy vaccinated subjects in this dataset.

>Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

There are 112 female and 60 male subjects/patients in this dataset.

> In terms of race and gender is this dataset representative of the US population? 

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```
Let's read some more database tables:

```{r}
specimen <- read_json("http://cmi-pb.org/api/v5_1/specimen",
                      simplifyVector = TRUE)

ab_titer <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer",
                      simplifyVector = TRUE)
```

```{r}
head(specimen)
```

```{r}
head(ab_titer)
```

```{r}
library(lubridate)
```

What is today’s date (at the time I am writing this obviously)

```{r}
today()
```

How many days have passed since new year 2000

```{r}
today() - ymd("2000-01-01")
```

What is this in years?

```{r}

```

To analyze this data we need to first "join" the different tables so we have all the data in one place, not spread across different tables. 

We can use the `*_join()` family of functions from **dplyr** to do this. 

```{r}
library(dplyr)

meta <- inner_join(subject, specimen)
head(meta)
```

```{r}
abdata <- inner_join(ab_titer, meta)
head(abdata)
```

> Q. What antibody isotypes are measured for these patients? 

```{r}
table(abdata$isotype)
```


> Q. What antigens are reported?

```{r}
table(abdata$antigen)
```

Let's focus on the IgG isotype and make a plot of MFI_normalized for all antigens. 

```{r}
igg <- abdata |> 
  filter(isotype == "IgG")

head(igg)
```

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot()
```

> Q. Is there a difference for aP vs wP individuals with these values? 

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() +
  facet_wrap(~infancy_vac)
```

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col = infancy_vac) +
  geom_boxplot()
```

> Q. Is there a temporal reponse - i.e. do these values increase or decease over time? 

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col = infancy_vac) +
  geom_boxplot() +
  facet_wrap(~visit)
```

## Focus on "PT" Pertussis Toxin antigen

```{r}
pt.igg.21 <- igg |> filter(antigen =="PT",
              dataset == "2021_dataset")
```


```{r}
ggplot(pt.igg.21) +
  aes(planned_day_relative_to_boost, 
      MFI_normalised, 
      col = infancy_vac,
      group = subject_id) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 14, lty =2)
```



>Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?










