---
title: "Class 17: Obtaining and processing SRA datasets on AWS"
author: "Cyrus Shabahang (PID:19145663)"
format: pdf
---

# Class 17 Lab

>Q1. What shell command can you use to view the top few lines of your FASTQ file? 

```bash
head SRR600956.fastq
```
>Q2. What length are these sequence reads? 

The sequence reads are 38 bases long. 

>Q3. Can you use the grep command to determine how many total reads are in this file?

grep -c "@SRR600956" SRR600956.fastq

This tells us 25849655 total reads. 

>Q4. Does your number of reads from grep match the name of the last read in the file? If not why not?

grep -c "@SRR600956" SRR600956.fastq

Yes, the number of reads from the grep matches the name of the last read in the file. 

>Q5. How would you check that the .fastq files actually look like what we expect for a FASTQ file?

head SRR2156848_1.fastq

>Q6. How could you check the number of sequences in each file?

grep -c "^@SRR2156848" SRR2156848_1.fastq

2959900 sequences in each file.

>Q7. Check your answer with the bottom of the file using tail and also check the matching mate pair FASTQ file. Do these numbers match? If so why or why not?

tail SRR2156848_1.fastq
tail SRR2156848_2.fastq

Yes, these number match. 

>Q8. Check you have pairs of FASTQ files for all four datasets and that they have the same number of counts in each pair. 

ls *.fastq
grep -c "^@SRR" *.fastq

>Q9. Fill in the missing kallisto installation/ setup commands

Unzip and untar: 
tar -zxvf kallisto_linux-v0.44.0.tar.gz

Add kallisto:
export PATH=$PATH:/home/ubuntu/kallisto_linux-v0.44.0

Print citation: 
kallisto cite

>Q10. Complete the remianing quantification commands

kallisto quant -i hg19.ensembl -o SRR2156849_quant SRR2156849_1.fastq SRR2156849_2.fastq
kallisto quant -i hg19.ensembl -o SRR2156850_quant SRR2156850_1.fastq SRR2156850_2.fastq
kallisto quant -i hg19.ensembl -o SRR2156851_quant SRR2156851_1.fastq SRR2156851_2.fastq

>Q11. Have a look at the TSV format versions of these files to understand their structure. What do you notice about these fils contents? 

These abudnance.tsv files all contain similar structure for each sample. Each file has the same transcript-level quantification results. 

```{r}
library(tximport)
library(ggplot2)
library(ggrepel)
library(rhdf5)
```

Kallisto results: 

```{r}
folders <- dir(pattern = "SRR21568.*_quant$")
samples <- sub("_quant", "", folders)
files <- file.path(folders, "abundance.h5")
names(files) <- samples

files
txi.kallisto <- tximport(files, type = "kallisto", txOut = TRUE)
head(txi.kallisto$counts)
```

Counting summaries:

```{r}
colSums(txi.kallisto$counts)
sum(rowSums(txi.kallisto$counts) > 0)
```

Now let's filter the transcripts:

```{r}
to.keep <- rowSums(txi.kallisto$counts) > 0
kset.nonzero <- txi.kallisto$counts[to.keep,]

keep2 <- apply(kset.nonzero, 1, sd) > 0
x <- kset.nonzero[keep2,]

dim(x)
```

#PCA

```{r}
pca <- prcomp(t(x), scale = TRUE)
summary(pca)
```

##PC1 vs PC2

```{r}
mycols <- c("blue", "blue", "red", "red")
ggplot(pca$x) +
  aes(PC1, PC2, label = rownames(pca$x)) +
  geom_point(col = mycols) +
  geom_text_repel(col = mycols) +
  theme_bw()
```

##PC1 vs PC3

```{r}
ggplot(pca$x) +
  aes(PC1, PC3, label = rownames(pca$x)) +
  geom_point(col = mycols) +
  geom_text_repel(col = mycols) +
  theme_bw()
```

##PC2 vs PC3

```{r}
ggplot(pca$x) +
  aes(PC2, PC3, label = rownames(pca$x)) +
  geom_point(col = mycols) +
  geom_text_repel(col = mycols) +
  theme_bw()
```


